---
title: "hw4"
format: pdf
editor: visual
---

## Q1

The debt data arise from a large postal survey on the psychology of debt. The frequency of credit card use is a three-level factor ranging from never, through occasionally to regularly.

### a

Declare the response as an ordered factor and make a plot showing the relationship to prodebt. Comment on the plot. Use a table or plot to display the relationship between the response and the income group.
```{r}
library(faraway)
library(ggplot2)
library(dplyr)

data(debt)

# Convert 'ccarduse' to an ordered factor
debt$ccarduse <- factor(debt$ccarduse, levels = c(1, 2, 3), labels = c("never", "occasionally", "regularly"), ordered = TRUE)

# Boxplot: prodebt vs. ccuse
ggplot(debt, aes(x = ccarduse, y = prodebt)) +
  geom_boxplot() +
  labs(
    x = "Credit-card use",
    y = "Pro-debt attitude",
    title = "Pro-debt Attitude by Credit-Card Use"
  )

# Stacked bar: ccuse by income
ggplot(debt, aes(x = incomegp, fill = ccarduse)) +
  geom_bar(position = "fill") +
  labs(
    x = "Income group",
    y = "Proportion",
    fill = "Card use",
    title = "Credit-Card Use Distribution across Income"
  ) +
    scale_y_continuous(labels = scales::percent_format())


# Table showing the relationship between 'ccarduse' and 'incomegp'
table(debt$ccarduse, debt$incomegp)
```
In the box plots, the median pro-debt score increases as card use increases. Individuals who regularly use credit cards tend to have higher pro-debt scores, indicating a more positive attitude toward debt which supports the idea that attitude toward debt is positively associated with credit card usage frequency.\
As income increases, the proportion of regular credit card users increases, and the proportion of non-users decreases. Income group 1 (lowest) has the highest proportion of “never” users, while income group 5 (highest) has the highest proportion of “regular” users. This suggests a positive association between income and frequency of credit card use.


### b
Fit a proportional odds model for credit card use with all the other variables as predictors. What are the two most significant predictors (largest t-values) and what is their qualitative effect on the response? What is the least significant predictor?
```{r}
library(MASS)

# Fit the proportional odds model
model_full <- polr(ccarduse ~ incomegp + house + children + singpar + agegp + bankacc + bsocacc + manage + cigbuy + xmasbuy + locintrn + prodebt, data = debt, Hess = TRUE, na.action = na.omit)

# Summary of the model
summary(model_full)

# Calculate p-values
ctable <- coef(summary(model_full))
p_values <- pnorm(abs(ctable[, "t value"]), lower.tail = FALSE) * 2
ctable <- cbind(ctable, "p value" = p_values)
ctable[order(abs(ctable[, "t value"]), decreasing = TRUE), ]
```
 The two most significant predictors were income group and having a bank account. Higher income levels were associated with greater odds of using credit cards more frequently. Similarly, individuals with a bank account were more likely to use credit cards. The least significant predictor was housing tenure, which showed no meaningful relationship with credit card use frequency in this model.
 

### c

Fit a proportional odds model using only the least significant predictor from the previous model. What is the significance of this predictor in this small model? Are the conclusions regarding this predictor contradictory for the two models?
```{r}
model_small <- polr(ccarduse ~ house, data = debt, Hess = TRUE, na.action = na.omit)

# Summary of the model
summary(model_small)

# Calculate p-value
ctab_c <- coef(summary(model_small))
pvals_c <- 2 * pnorm(-abs(ctab_c[, "t value"]))
cbind(ctab_c, `p value` = pvals_c)
```
The effect of housing becomes statistically significant in the small model, with a positive coefficient indicating that more secure housing is associated with more frequent credit card use. This finding contrasts with the full model, where housing tenure was the least significant predictor and not statistically different.
The contradiction could be because, in the full model, the effect of housing may have been masked by collinearity or shared variance with other predictors like income or bank access. But in isolation, housing tenure shows a clear association with card use.

### d

Use stepwise AIC to select a smaller model than the full set of predictors. You will need to handle the missing values carefully. Report on the qualitative effect of the predictors in your chosen model. Can we conclude that the predictors that were dropped from the model have no relation to the response?
```{r}
library(mice)

# Select relevant columns
vars <- c("ccarduse", "incomegp", "house", "children", "singpar", 
          "agegp", "bankacc", "bsocacc", "manage", "cigbuy", 
          "xmasbuy", "locintrn", "prodebt")

# Subset the data
debt_sub <- debt[, vars]

# Impute missing data using mice
set.seed(123)
imputed_data <- mice(debt_sub, m = 1, method = "pmm", maxit = 10)

debt_imputed <- complete(imputed_data)

# Fit model and run stepwise selection
model_full_imputed <- polr(ccarduse ~ ., data = debt_imputed, Hess = TRUE)
model_step_imputed <- stepAIC(model_full_imputed, direction = "both", trace = FALSE)
summary(model_step_imputed)
```
The final model retained six predictors: income group, age group, bank account status, ability to manage finances, cigarette buying, and pro-debt attitude. Higher income, older age, having a bank account, and stronger pro-debt attitudes were associated with more frequent credit card use. In contrast, cigarette buyers were less likely to use credit cards. Although some predictors were dropped. This does not imply they have no relationship with the response. It only means that they did not improve model fit beyond the retained variables.


### e

Compute the median values of the predictors in your selected model. At these median values, contrast the predicted outcome probabilities for both smokers and nonsmokers.
```{r}
library(dplyr)

med_values <- tibble(
  incomegp = median(debt_imputed$incomegp, na.rm = TRUE),
  agegp    = median(debt_imputed$agegp,    na.rm = TRUE),
  bankacc  = as.numeric(names(which.max(table(debt_imputed$bankacc)))),
  manage   = median(debt_imputed$manage,   na.rm = TRUE),
  prodebt  = median(debt_imputed$prodebt,  na.rm = TRUE)
)

new_data <- bind_rows(
  mutate(med_values, cigbuy = 1),
  mutate(med_values, cigbuy = 0)
)

# Predict
predicted_probs <- predict(model_step_imputed, newdata = new_data, type = "probs")
rownames(predicted_probs) <- c("Smoker (cigbuy=1)", "Non-smoker (cigbuy=0)")
print(round(predicted_probs, 4))
```


### f

Fit a proportional hazards model to the same set of predictors and recompute
the two sets of probabilities from the previous question. Does it make a differ-
ence to use this type of model?

## Q5

Consider the Galápagos data and model analyzed in this chapter. The purpose of this question is to reproduce the details of the GLM fitting of this data.

### a

Fit a Poisson model to the species response with the five geographic variables as predictors. Do not use the endemics variable. Report the values of the coefficients and the deviance.

### b

For a Poisson GLM, derive n, dn/du, Vu) and the weights to be used in an iteratively fit GLM. What is the form of the adjusted dependent variable here?

### c

Using the observed response as initial values, compute the first stage of the iteration, stopping after the first linear model fit. Compare the coefficients of this linear model to those found in the GLM fit. How close are they?

### d

Continue the iteration to get the next n and u. Use this to compute the current value of the deviance. How close is this to the deviance from the GLM?

### e

Compute one more iteration of the GLM fit, reporting the next calculation of the coefficients and deviance. How close are these to target now?

### f

Repeat these iterations a few more times, computing the deviance in each time. Stop when the deviance does not change much. Compare your final estimated coefficients to that produced by the GLM fit.

### g

Use your final iterated linear model fit to produce standard errors for the coef-ficients. How close are these to that produced by the direct GLM fit?

## Q6

Again using the Galápagos data, fit a Poisson model to the species response with the five geographic variables as predictors. Do not use the endemics variable. The purpose of this question is to compare six different ways of testing the significance of the elevation predictor, i.e., Ho : BElev = 0. In each case, report the p-value.

### a

Use the z-statistic from the model summary.

### b

Fit a model without elevation and use the difference in deviances to make the test.

### c

Use the Pearson Chi-squared statistic in place of the deviance in the previous test.

### d

Fit the Poisson model with a free dispersion parameter as described in Section 5.2. Make the test using the model summary.

### e

Use the sandwich estimation method for the standard errors in the original model. Use these to compute z-statistics.

### f

Use the robust GLM estimation method and report the test result from the summary.

### g

Compare all six results. Pick the best one and justify your choice.
