---
title: "hw5"
format: pdf
editor: visual
---

## Q3
The ratdrink data consist of five weekly measurements of body weight for 27 rats. The first 10 rats are on a control treatment while 7 rats have thyroxine added to their drinking water. Ten rats have thiouracil added to their water.\


### a 
Plot the data showing how weight increases with age on a single panel, taking care to distinguish the three treatment groups. Now create a three-panel plot, one for each group. Discuss what can be seen.
```{r}
library(faraway)
library(dplyr)
data(ratdrink)
glimpse(ratdrink)
summary(ratdrink)
library(ggplot2)

# One panel plot
ggplot(ratdrink, aes(x = weeks, y = wt, group = subject, color = treat)) +
  geom_line(alpha = 0.6) +
  labs(title = "Rat Weight Over Time by Treatment",
       x = "Weeks", y = "Weight (g)", color = "Treatment") +
  theme_minimal()

# Three-panel faceted plot by treatment
ggplot(ratdrink, aes(x = weeks, y = wt, group = subject)) +
  geom_line(aes(color = subject), show.legend = FALSE) +
  facet_wrap(~ treat) +
  labs(title = "Weight Trajectories by Treatment Group",
       x = "Weeks", y = "Weight (g)") +
  theme_minimal()
```
From the single panel plot we can see that all groups tend to show increasing weight over time. The second (three-panel) plot shows the weight of rats over time, with different treatments separated via the panels. The three-panel plot allows for a clearer comparison of how each treatment group (control, thyroxine, thiouracil) affects weight gain over the weeks. The control group shows a steady increase in weight, while the thyroxine group appears to gain weight slightly more rapidly. The thiouracil group shows a slower increase, indicating a potential suppression of weight gain compared to the other groups.

### b 
Fit a linear longitudinal model that allows for a random slope and intercept for each rat. Each group should have a different mean line. Give interpretation for the following estimates:
i. The fixed effect intercept term.\
ii. The interaction between thiouracil and week.\
iii. The intercept random effect SD.\
```{r}
library(lme4)

model_b <- lmer(wt ~ weeks * treat + (weeks | subject), data = ratdrink)

summary(model_b)
```
Interpretation of the estimates:\
i. The fixed effect intercept term represents the average weight of a rat at week 0 (the baseline) across all treatment groups. It serves as the starting point for the weight trajectory of the rats before any treatment effects are applied. At week 0, rats in the control group weigh an average of 52.88 grams.\

ii. The interaction between thiouracil and week indicates how the effect of time (weeks) on weight differs for rats receiving thiouracil compared to the control group. A negative coefficient suggests that as weeks progress, the weight gain in the thiouracil group is less than that in the control group, indicating a potential suppression of weight gain due to thiouracil treatment. Rats in the thiouracil group gain weight at a slower rate than control rats by about 9.37 grams less per week on average, (after adjusting for random individual variation and initial group differences). Even though the thiouracil group has a positive main effect (+4.78), this interaction shows that their growth rate over time is significantly lower than control.\

iii. The intercept random effect SD represents the variability in the initial weights of the rats across different subjects. A higher value indicates that there is considerable individual variation in starting weights among the rats, suggesting that some rats start at a significantly different weight than others, regardless of treatment. In this case, individual rats vary around the group-level intercept (52.88g) with a standard deviation of about 5.7 grams. That is, some rats start heavier or lighter than average, even within the same treatment group.\

### c 
Check whether there is a significant treatment effect.
```{r}
# Full model (same as model_b)
model_full <- lmer(wt ~ weeks * treat + (weeks | subject), data = ratdrink)

# Reduced model (no treatment terms)
model_reduced <- lmer(wt ~ weeks + (weeks | subject), data = ratdrink)

# Likelihood ratio test
anova(model_reduced, model_full)
```
The likelihood ratio test comparing the reduced model (only time effect) to the full model (which includes treatment and treatment-by-time interactions) yields a Chi-squared statistic of 25.14 on 4 degrees of freedom, with a p-value of 4.723e-05. This provides strong evidence that treatment has a significant effect on weight and including treatment in the model improves the model fit significantly.

### d 
Construct diagnostic plots showing the residuals against the fitted values and a QQ plot of the residuals. Interpret.
```{r}
resid <- residuals(model_full)
fitted_vals <- fitted(model_full)

# Residuals vs Fitted Plot
plot(fitted_vals, resid,
     xlab = "Fitted Values", ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red")

# QQ Plot
qqnorm(resid, main = "Normal Q-Q Plot of Residuals")
qqline(resid, col = "red")
```
The residuals vs. fitted values plot shows no clear pattern, indicating that the model assumptions of homoscedasticity (constant variance) and linearity are reasonably met. The residuals are randomly scattered around zero, suggesting that the model fits the data well without systematic errors. The QQ plot shows that the residuals mostly follow a straight line, indicating that the assumption of normality of residuals is also reasonably satisfied, though there are very minor deviations at the tails.


### e 
Construct confidence intervals for the parameters of the model. Which random effect terms may not be significant? Is the thyroxine group significantly different from the control group?


## Q4
The ohio data concern 536 children from Steubenville, Ohio and were taken as part of a study on the effects of air pollution. Children were in the study for 4 years from ages 7 to 10. The response was whether they wheezed or not. The variables are:\
resp: an indicator of wheeze status (1 = yes, 0 = no)\
id: an identifier for the child\
age: 7 yrs = -2, 8 yrs = - 1, 9 yrs = 0, 10 yrs = 1\
smoke: an indicator of maternal smoking at the first year of the study (1 = smoker, 0 = nonsmoker)\


### a
Do any of the mothers in the study change their smoking status during the period of observation?


### b 
Construct a table that shows proportion of children who wheeze for 0, 1, 2, 3 or 4 years broken down by maternal smoking status.


### c 
Make plot which shows how the proportion of children wheezing changes by age with a separate line for smoking and nonsmoking mothers.


### d 
Group the data by child to count the total (out of four) years of wheezing.\
Fit a binomial GLM to this response to check for a maternal smoking effect.\
Does this prove there is a smoking effect or could there be another plausible explanation?\


### e 
Fit a model for each individual response using a GLMM fit using penalized quasi-likelihood. Describe the effects of age and maternal smoking. How do the odds of wheezing change numerically over time?


### f 
Now fit the same model but using adaptive Gaussian-Hermit quadrature. Compare to the previous model fit.


### g 
Use INLA to fit the same model. What does this model say about the effect of age and maternal smoking?


### h 
Use STAN to fit the same model. Check the MCMC diagnostics and again discuss the age
and maternal smoking effects.


### i
Fit the model using GEE. Use an autoregressive rather than exchangeable error structure. Compare the results to the previous model fits. In your model, what indicates that a child who already wheezes is likely to continue to wheeze?


### j 
What is your overall conclusion regarding the effect of age and maternal smok-ing? Can we trust the GLM result or are the GLMM models preferable?